---
title: 'Statistical Tests for Categorical Data (part II)'
output: 
  html_document:
    code_folding: show 
    df_print: paged
    toc: true
    toc_float:
      collapsed: false
    number_sections: false
    theme: spacelab
    highlight: tango
    includes:
      after_body: ../footerEA.html
    css: ../style.css
---


```{r setup, include=FALSE}
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```


```{r, echo = FALSE, purl = FALSE}
knitr::knit_hooks$set(purl = knitr::hook_purl)

options(purl = FALSE)

knitr::opts_chunk$set(purl = FALSE)
```

```{r, include = FALSE}
knitr::opts_hooks$set(eval = function(opt) {
  if (any(opt$exercise))
    opt$eval <- opt$include <- FALSE
  
  opt
})


static <- TRUE

options(width = 100)

```


```{r packages, include = FALSE}
library(kableExtra)
library(knitr)
```


## Preface {data-progressive=FALSE}

Open Rstudio to do the practicals. Note that tasks with * are optional.

### R Packages

In this practical, a number of R packages are used.
The packages used (with versions that were used to generate the solutions) are:

* `survival` (version: `r packageVersion("survival")`)

_`r R.version.string`_

### Two sample tests (paired data) {.tabset .tabset-fade .tabset-pills}

```{r, eval = static, echo = FALSE}
asis_output("#### Task 1\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#0c2074">')
```

Explore the help page of the R function `mcnemar.test()`.

* Use the code from the example to obtain the matrix `Performance`.
* We want to investigate whether the percentage of approve is different between the 2 surveys (which are taken with one month apart). Define the null and alternative hypothesis. Let's assume that the samples are dependent.
* Perform the test and save the results in an object called test_res_cat_paired. What is the conclusion?
* Calculate the test statistic manually and confirm the result with the R function.



```{r onesample1b_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#onesample1b">Hint</button>
<div id = "onesample1b" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#6d79ac">
Use the R function mcnemar.test() to investigate whether the percentages of the two groups are different in a matched setting.
</div>')
```





```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 1\\n")
```

```{r onesample1-solution, solution = TRUE}
Performance <-
matrix(c(794, 86, 150, 570),
       nrow = 2,
       dimnames = list("1st Survey" = c("Approve", "Disapprove"),
                       "2nd Survey" = c("Approve", "Disapprove")))
Performance
```

Null hypothesis: the percentage of approve in the 1st survey is the same with the percentage of approve in the 2nd survey.\
Alternative hypothesis: the percentage of approve in the 1st survey is different compared to the percentage of approve in the 2nd survey.



```{r onesample1-solutionb, solution = TRUE}
test_res_cat_paired <- mcnemar.test(Performance)
test_res_cat_paired
```

If we assume that the significant level is equal to 0.05, the output indicates that we can reject the null hypothesis that the percentages are the same.


The test statistic (with continuity correction) is: $X^2=\frac{(\mid b-c \mid -1)^2}{b+c}$,\
where\

```{r, echo = FALSE}
set.seed(2021)
x1 <- sample(0:1, 100, replace = TRUE)
x2 <- sample(0:1, 100, replace = TRUE)

tab <- table(x1, x2)

tab <- matrix(c("a", "b", "a + b", "c", "d", "c + d", "a + c", "b + d", "n"), 3, 3, byrow = TRUE)

colnames(tab) <- c("Approve (2st Survey)", "Disapprove (2st Survey)", "Total")
rownames(tab) <- c("Approve (1st Survey)", "Disapprove (1st Survey)", "Total")

library(knitr)
kable(tab)
```

```{r onesample1-solutionc, solution = TRUE}
test_statistic <- (abs(150-86)-1)^2/(150+86)
test_statistic
```















```{r, eval = static, echo = FALSE}
asis_output("#### Task 2*\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#0c2074">')
```

* Create the following data:
     - `id`: a vector that consist of the numbers 1 until 100 (integers) where each number is repeated 2 times
     - `case`: a vector that repeats `c(0, 1)` 100 times
     - `response`: a vector of length 200 with randomly selected 0 and 1 values
* Note that the data is in the long format. Use the code `table(response[case == 0], response[case == 1])` to generate the paired table.
* Investigate whether the percentage of response is different between cases and controls (defined by the case variable). Let's assume that the samples are dependent. What is the conclusion?
* Investigate the same hypothesis using a conditional logistic regression.



```{r onesample2b_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#onesample2b">Hint</button>
<div id = "onesample2b" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#6d79ac">
Use the R function table() to create the paired table.
</div>')
```



```{r onesample2a_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#onesample2a">Hint</button>
<div id = "onesample2a" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#6d79ac">
Use the R function mcnemar.test() to investigate whether the percentages of the two groups are different in a matched setting.
</div>')
```


```{r onesample2c_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#onesample2c">Hint</button>
<div id = "onesample2c" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#6d79ac">
Use the R function clogit() from the survival package to perform a conditional logistic regression.
</div>')
```



```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 2*\\n")
```

```{r onesample2-solution, solution = TRUE}
# Generate the data
set.seed(123)
n = 100
id <- rep(x = 1:n, each = 2)
case <- rep(x = 0:1, times = n)
response <- rbinom(n = n*2, size = 1, prob = 0.5)

# Create the paired table
table(response[case == 0], response[case == 1])

# Perform the McNemar test
mcnemar.test(table(response[case == 0], response[case == 1]))
```

If we assume that the significant level is equal to 0.05, the output indicates that we cannot reject the null hypothesis that the percentages are the same.

```{r onesample2-solutionb, solution = TRUE}
# Fit the conditional logistic model
library(survival)
summary(clogit(response ~ case + strata(id)))
```

The p-values are similar. You can repeat the analysis assuming a different set.seed().





















### Multiple testing {.tabset .tabset-fade .tabset-pills}

```{r, eval = static, echo = FALSE}
asis_output("#### Task 1\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#0c2074">')
```


* Create the following data:
    - before: a vector with length 500 that takes values 0 and 1.
    - after: a vector with length 500 that takes values 0 and 1.
    - sex: a vector with length 500 that consists of males and females.
* Create the paired table for males. Investigate whether the percentage of 1s is different between before and after in male patients. Let’s assume that the samples (before/after) are dependent. Save the results in an object called test_males.
* Create the paired table for females. Investigate whether the percentage of 1s is different between before and after in female patients. Save the results in an object called test_females.
* Extract the p-values from the tests and save them to the objects called p_males and p_females. Correct the p-values for multiple testing using the bonferroni method.

```{r multitest1_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#multitest1">Hint</button>
<div id = "multitest1" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#6d79ac">
Use the R function table() to create the paired table.
</div>')
```

```{r multitest1a_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#multitest1a">Hint</button>
<div id = "multitest1a" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#6d79ac">
Use the R function mcnemar.test() to investigate whether the percentages of the two groups are different in a matched setting.
</div>')
```

```{r multitest1b_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#multitest1b">Hint</button>
<div id = "multitest1b" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#6d79ac">
Use the R function p.adjust() to return p-values adjusted for multiple testing.
</div>')
```



```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 1\\n")
```



```{r multitest1-solution, solution = TRUE}
# Create the data:
set.seed(123)
before <- sample(x = c(0, 1), size = 500, replace = TRUE)
after <- sample(x = c(0, 1), size = 500, replace = TRUE)
sex <- sample(x = c("males", "females"), size = 500, replace = TRUE)

dat <- data.frame(before, after, sex)

# Create the paired table for males:
tab_males <- table(dat$before[dat$sex == "males"], dat$after[dat$sex == "males"])

# Perform the test for males:
test_males <- mcnemar.test(tab_males)

# Create the paired table for females:
tab_females <- table(dat$before[dat$sex == "females"], dat$after[dat$sex == "females"])

# Perform the test for females :
test_females <- mcnemar.test(tab_females)

# Extract the p-value from the tests:
p_males <- test_males$p.value
p_females <- test_females$p.value

# Apply the correction for multiple testing:
p.adjust(p = c(p_males, p_females), method = "bonferroni")
         
# Alternatively  
m <- 2
pmin(c(p_males, p_females) * m, 1)
```











```{r, eval = static, echo = FALSE}
asis_output("#### Task 2*\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#0c2074">')
```

Explore the pre-loaded data set `iris`. 

* Investigate whether the mean sepal length is different between:
    - `setosa` and `versicolor` 
    - `setosa` and `virginica`
    - `versicolor` and `virginica`
* Correct the p-values for multiple testing using the holm method.


```{r multitest2_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#multitest2">Hint</button>
<div id = "multitest2" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#6d79ac">
Since the data looks normally distributed and we have 50 subjects, we can use the t-test. Use the R function t.test() to investigate whether the mean values of two sample groups are different.
</div>')
```

```{r multitest2a_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#multitest2a">Hint</button>
<div id = "multitest2a" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#6d79ac">
Use the R function p.adjust() to return p-values adjusted for multiple testing.
</div>')
```



```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 2*\\n")
```



```{r multitest2-solution, solution = TRUE}
# Explore the data:
head(iris)
boxplot(iris$Sepal.Length ~ iris$Species)

par(mfrow = c(1, 3))
hist(iris$Sepal.Length[iris$Species == "setosa"], 
     main = "setosa", xlab = "Length")
hist(iris$Sepal.Length[iris$Species == "versicolor"], 
     main = "versicolor", xlab = "Length")
hist(iris$Sepal.Length[iris$Species == "virginica"], 
     main = "virginica", xlab = "Length")


# Perform the tests:
test1 <- t.test(x = iris$Sepal.Length[iris$Species == "setosa"], 
                y = iris$Sepal.Length[iris$Species == "versicolor"])
test2 <- t.test(x = iris$Sepal.Length[iris$Species == "setosa"], 
                y = iris$Sepal.Length[iris$Species == "virginica"])
test3 <- t.test(x = iris$Sepal.Length[iris$Species == "versicolor"], 
                y = iris$Sepal.Length[iris$Species == "virginica"])

# Apply the correction for multiple testing:
p.adjust(p = c(test1$p.value, test2$p.value, test3$p.value), method = "holm")

# Alternatively
p <- c(test1$p.value, test2$p.value, test3$p.value)

# order the p-values from the smallest to the largest:
p_order <- p[order(p)]

# set the number of tests:
m <- 3

# obtain the 1st adjusted p-value:
i <- 1
p_adjust_1 <- (m-i+1) * p_order[i]

# obtain the 2st adjusted p-value:
i <- 2
p_adjust_2 <- (m-i+1) * p_order[i]

# obtain the 3st adjusted p-value:
i <- 3
p_adjust_3 <- (m-i+1) * p_order[i]

# bring the adjusted p-values in the correct order:
c(p_adjust_1, p_adjust_2, p_adjust_3)[order(p)]
```

