---
title: "Biostatistics I: Descriptive Statistics"
subtitle: "Correlation"
author: "Eleni-Rosalina Andrinopoulou"
institute: "Department of Biostatistics, Erasmus Medical Center"
email: "e.andrinopoulou@erasmusmc.nl"
twitter: "@erandrinopoulou"
output:
  beamer_presentation: 
    template: mytemplate.latex
    includes:
      in_header: SlideTemplate.tex
    keep_tex: yes
    incremental: false
classoption: aspectratio=169
---

  
```{r setup, include=FALSE}
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\n \\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})


knitr::knit_hooks$set(
  nospace = function(before, options, envir) {
    if (before) {
      knitr::asis_output("\\vspace*{-1.5ex}")
    }
  }
)

knitr::opts_chunk$set(echo = FALSE, comment=NA)

library(survival)
library(knitr)
library(kableExtra)
library(dplyr)

pbc <- survival::pbc
pbcseq <- survival::pbcseq
```


## In this Section


* Correlation coefficients

* Examples



## Correlation

Correlation is a measure that describes the strength of the association between two variables. Let’s assume that we have two continuous variables, we can get the following relationships:
\newline
\begincols
  \begincol{0.3\textwidth}

\blue{Positive correlation}

```{r, echo = FALSE, cache = TRUE, message = FALSE, warning = FALSE, fig.height = 4, fig.width = 4}
library(MASS)
set.seed(2021+1)

r = 0.83
data = mvrnorm(n = 50, mu = c(0, 0), 
               Sigma = matrix(c(1, r, r, 1), nrow=2), 
               empirical = TRUE)
x = data[, 1]  
y = data[, 2]  
dat <- data.frame(x = x, y = y)

plot(x, y, pch = 16, xlab = "Variable 1", ylab = "Variable 2")
```

 \endcol
\begincol{0.3\linewidth}

\blue{Negative correlation}

```{r, echo = FALSE, cache = TRUE, message = FALSE, warning = FALSE, fig.height = 4, fig.width = 4}
library(MASS)
set.seed(2021+1)

r = -0.83
data = mvrnorm(n = 50, mu = c(0, 0), 
               Sigma = matrix(c(1, r, r, 1), nrow=2), 
               empirical = TRUE)
x = data[, 1]  
y = data[, 2]  
dat <- data.frame(x = x, y = y)

plot(x, y, pch = 16, xlab = "Variable 1", ylab = "Variable 2")
```

 \endcol
\begincol{0.3\linewidth}

\blue{No correlation}

```{r, echo = FALSE, cache = TRUE, message = FALSE, warning = FALSE, fig.height = 4, fig.width = 4}
library(MASS)
set.seed(2021+1)

r = 0.2
data = mvrnorm(n = 50, mu = c(0, 0), 
               Sigma = matrix(c(1, r, r, 1), nrow=2), 
               empirical = TRUE)
x = data[, 1]  
y = data[, 2]  
dat <- data.frame(x = x, y = y)

plot(x, y, pch = 16, xlab = "Variable 1", ylab = "Variable 2")
```     

 \endcol
\endcols


## Pearson Correlation

- magnitude of association
- linear association
- direction of the relationship

A relationship is linear when a change in one variable is associated with a proportional change in the other variable

Pearson Correlation: $corr(X, Y) = \frac{cov(X, Y)}{sd(X)sd(Y)}$,

where $cov(X, Y)$ is the covariance and $sd(X)$, $sd(Y)$ are the standard deviations


## Spearman Correlation

- direction of the relationship
- monotonic relationship

In a monotonic relationship, the variables tend to change together, but not always at a constant rate (as in the linear case)

The Spearman correlation coefficient is based on the ranked values:
$corr_R(X, Y) = \frac{cov(R_X, R_Y)}{sd(R_X)sd(R_Y)}$
\newline

\blue{\textit{What is rank?}}
Ranks are integers indicating the rank of some values. E.g. the rank of 3, 10, 16, 6, 2 is 2, 4, 5, 3, 1:
```{r, echo = TRUE, cache = TRUE, message = FALSE, warning = FALSE, comment = ""}
rank(c(3, 10, 16, 6, 2))
```


## Difference between Pearson and Spearman

\begincols
  \begincol{0.5\textwidth}

\blue{Weak positive correlation:}
      
```{r, echo = FALSE, cache = TRUE, message = FALSE, warning = FALSE, fig.height = 4, fig.width = 4}
library(MASS)
r = 0.5
set.seed(123)
add_legend <- function(...) {
  opar <- par(fig=c(0, 1, 0, 1), oma=c(0, 0, 0, 0), 
    mar=c(0, 0, 0, 0), new=TRUE)
  on.exit(par(opar))
  plot(0, 0, type='n', bty='n', xaxt='n', yaxt='n')
  legend(...)
}
data = mvrnorm(n = 50, mu = c(0, 0), 
               Sigma = matrix(c(1, r, r, 1), nrow=2), 
               empirical = TRUE)
x = data[, 1]  
y = data[, 2]  
dat <- data.frame(x = x, y = y)

par(mar = c(5, 4, 1.7, 0.2))
plot(x, y, pch = 16, xlab = "Variable 1", ylab = "Variable 2")
p1 <- cor(x, y, method = "pearson")
p2 <- cor(x, y, method = "spearman")
add_legend("topright", c("Pearson = 0.5", "Spearman = 0.51"), horiz = TRUE, bty = 'n') 
```         


 \endcol
\begincol{0.5\linewidth}

\blue{Strong positive correlation:}

```{r, echo = FALSE, cache = TRUE, message = FALSE, warning = FALSE, fig.height = 4, fig.width = 4}
library(MASS)
set.seed(123)
add_legend <- function(...) {
  opar <- par(fig=c(0, 1, 0, 1), oma=c(0, 0, 0, 0), 
    mar=c(0, 0, 0, 0), new=TRUE)
  on.exit(par(opar))
  plot(0, 0, type='n', bty='n', xaxt='n', yaxt='n')
  legend(...)
}

r = 0.9
data = mvrnorm(n = 50, mu = c(0, 0), 
               Sigma = matrix(c(1, r, r, 1), nrow=2), 
               empirical = TRUE)
x = data[, 1]  
y = data[, 2]  
dat <- data.frame(x = x, y = y)

par(mar = c(5, 4, 1.7, 0.2))
plot(x, y, pch = 16, xlab = "Variable 1", ylab = "Variable 2")
p1 <- cor(x, y, method = "pearson")
p2 <- cor(x, y, method = "spearman")
add_legend("topright", c("Pearson = 0.9", "Spearman = 0.8"), horiz = TRUE, bty = 'n')    
```   

 \endcol
\endcols



## Difference between Pearson and Spearman

\begincols
  \begincol{0.5\textwidth}

\blue{Linear VS monotonic relationship:}

```{r, echo = FALSE, cache = TRUE, message = FALSE, warning = FALSE, fig.height = 4, fig.width = 4}
set.seed(123)
add_legend <- function(...) {
  opar <- par(fig=c(0, 1, 0, 1), oma=c(0, 0, 0, 0), 
    mar=c(0, 0, 0, 0), new=TRUE)
  on.exit(par(opar))
  plot(0, 0, type='n', bty='n', xaxt='n', yaxt='n')
  legend(...)
}
x <- c(1:20)
y <- c(50, 50, 49.5, 49, 48.6, 48.3, 47.7, 47, 46.6, 46.4, 45.9, 45.5, 45, 40, 35, 25, 20, 10, 5, 0)


par(mar = c(5, 4, 1.7, 0.2))
plot(x, y, pch = 16, xlab = "Variable 1", ylab = "Variable 2")
p1 <- cor(x, y, method = "pearson")
p2 <- cor(x, y, method = "spearman")
add_legend("topright", c("Pearson = -0.85", "Spearman = -0.99"), horiz = TRUE, bty = 'n')
```  

 \endcol
\begincol{0.5\linewidth}

\blue{What if there is a correlation but this is not linear?}

```{r, echo = FALSE, cache = TRUE, message = FALSE, warning = FALSE, fig.height = 4, fig.width = 4}
set.seed(123)
add_legend <- function(...) {
  opar <- par(fig=c(0, 1, 0, 1), oma=c(0, 0, 0, 0), 
    mar=c(0, 0, 0, 0), new=TRUE)
  on.exit(par(opar))
  plot(0, 0, type='n', bty='n', xaxt='n', yaxt='n')
  legend(...)
}
x <- c(1:29)
y <- c(22, 20, 27, 25, 27, 38, 30, 33, 44, 47, 46, 49, 47, 48.3, 47.7, 47, 45, 46.4, 45.9, 45.5, 30, 45, 40, 34, 35, 32, 25, 20, 22)


par(mar = c(5, 4, 1.7, 0.2))
plot(x, y, pch = 16, xlab = "Variable 1", ylab = "Variable 2")
p1 <- cor(x, y, method = "pearson")
p2 <- cor(x, y, method = "spearman")
add_legend("topright", c("Pearson = 0.06", "Spearman = 0.04"), horiz = TRUE, bty = 'n')
``` 

 \endcol
\endcols


## Other Correlation Measures

  - \blue{Point-Biserial}: It evaluates the association between a continuous variable with a categorical (dichotomous) variable
  - \blue{Intraclass}: It evaluates the association between two continuous variables that are structured in groups 
  
\blue{\textit{Note}}

- Correlation must not be confused with causality
- If two variables are correlated, it does not imply that one variable causes the changes in another variable


