---
title: "Biostatistics I: Hypothesis testing"
subtitle: "Categorical data: Proportion tests"
author: "Eleni-Rosalina Andrinopoulou"
institute: "Department of Biostatistics, Erasmus Medical Center"
email: "e.andrinopoulou@erasmusmc.nl"
twitter: "@erandrinopoulou"
output:
  beamer_presentation: 
    template: mytemplate.latex
    includes:
      in_header: SlideTemplate.tex
    keep_tex: yes
    incremental: false
classoption: aspectratio=169
---

  
```{r setup, include=FALSE}
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\n \\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})


knitr::knit_hooks$set(
  nospace = function(before, options, envir) {
    if (before) {
      knitr::asis_output("\\vspace*{-1.5ex}")
    }
  }
)

knitr::opts_chunk$set(echo = FALSE, comment=NA)

library(survival)
library(knitr)
library(kableExtra)
library(dplyr)

pbc <- survival::pbc
pbcseq <- survival::pbcseq
```


## In this Section


* $z$-test for proportions

* Bionomial test

* Examples


 
## $z$-test for proportions: Theory   

\blue{Assumptions}

- The observations are independent of one another
- The sample size is large enough to use the normal approximation
$\mathcal{N}(n p, n p (1-p))$\
    - $np > 10$ and $n(1-p) > 10$, where $n$ is the number of observations and $p$ the proportion
    


## One sample $z$-test for proportions: Theory   

\blue{Scenario} 

Is the probability of being diagnosed with asthma now different than it was 50 years ago?

\blue{Hypothesis}

$H_0: \pi = \pi_0$\newline
$H_1: \pi \neq \pi_0$



## One sample $z$-test for proportions: Theory   

\blue{Hypothesis}

\begin{tcolorbox}[colback=EMCdark,
                  colframe=EMCdark,
                  width=13cm
                 ]
{\color{white} 
If \textbf{one-tailed}\newline
Is the probability of being diagnosed with asthma now higher than it was 50 years ago?\newline
$H_0:\pi=\pi_0$\newline
$H_1:\pi>\pi_0$\newline
\newline
or\newline
\newline
Is the probability of being diagnosed with asthma now lower than it was 50 years ago?\newline
$H_0:\pi=\pi_0$\newline
$H_1:\pi<\pi_0$
}

\end{tcolorbox}


## One sample $z$-test for proportions: Theory  

\blue{Test statistic}

For large sample sizes, the distribution of the test statistic is approximately normal

$z = \frac{p-\pi_0}{\sqrt{\frac{\pi_0(1-\pi_0)}{n}}}$

- Sample proportion: $p$
- Population proportion: $\pi_0$
- Number of subjects: $n$


If continuity correction is applied: $z = \frac{p-\pi_0 + c}{\sqrt{\frac{\pi_0(1-\pi_0)}{n}}}$, \

where\

- $c = -\frac{1}{2n}$ if $p > \pi_0$
- $c = \frac{1}{2n}$ if $p < \pi_0$
- $c =0$ if $|p-\pi_0| < \frac{1}{2n}$


## One sample $z$-test for proportions: Theory  

\blue{Sampling distribution}

- $z$-distribution 
- Critical values and p-value


\blue{Type I error}

- Normally $\alpha = 0.05$

\blue{Draw conclusions}

- Compare test statistic ($z$) with the critical values$_{\alpha/2}$ or the p-value with $\alpha$

\begin{tcolorbox}[colback=EMCdark,
                  colframe=EMCdark,
                  width=13cm
                 ]
{\color{white} 

If \textbf{one-tailed}: Compare test statistic with the critical value$_{\alpha}$

}

\end{tcolorbox}


## One sample $z$-test for proportions: Application

\blue{Scenario} 

Is the probability of being diagnosed with asthma now different than it was 50 years ago?

\blue{Hypothesis}

$H_0: \pi = \pi_0$\newline
$H_1: \pi \neq \pi_0$


## One sample $z$-test for proportions: Application

\begincols
  \begincol{0.4\textwidth}

\blue{Hypothesis}

$H_0: \pi = \pi_0$\newline
$H_1: \pi \neq \pi_0$

\vspace*{0.5cm}

\blue{Collect and visualize data}
\vspace*{0.2cm}
```{r, echo = FALSE, cache = TRUE, message = FALSE, warning = FALSE}
set.seed(2021)
x <- sample(c("No", "Yes"), 100, replace = TRUE)

kable(table(x))
```

50 years ago we had $\pi_0 = 0.6$

\endcol
\begincol{0.5\linewidth}

\blue{Test statistic}

(with no continuity correction):
$z = \frac{p-\pi_0}{\sqrt{\frac{\pi_0(1-\pi_0)}{n}}} = \frac{0.53-0.6}{\sqrt{\frac{0.6(1-0.6)}{100}}}=-1.43$

\vspace*{0.5cm}

\blue{Type I error}

$\alpha = 0.05$

 \endcol
\endcols


## One sample $z$-test for proportions: Application 

\blue{Critical values}

\begincols
  \begincol{0.6\linewidth}

Using ```R``` we get the critical values from the $z$-distribution:\newline
critical value$_{\alpha/2}$ = critical value$_{0.05/2}$

```{r, echo = TRUE, cache = TRUE, message = FALSE, warning = FALSE, comment=""}
qnorm(p = 0.05/2, lower.tail = FALSE)
```

-critical value$_{\alpha/2}$ = -critical value$_{0.05/2}$

```{r, echo = TRUE, cache = TRUE, message = FALSE, warning = FALSE, comment=""}
qnorm(p = 0.05/2, lower.tail = TRUE)
```

 \endcol
\begincol{0.4\linewidth}

\textcolor{white}{texttestestesstdfgzdfgdfgdfgdsdfg}
```{r, echo = FALSE, cache = TRUE, message = FALSE, warning = FALSE, fig.height=2.5, fig.width=4}
library(ggplot2)
mu = 0
SD = 1
x <- rt(1000000, df = 5)
RI <- c(mu - 2 * SD, mu + 2 * SD)
meanX <- mean(x)

dat <- with(density(x), data.frame(x, y))
ggplot(data = dat, mapping = aes(x = x, y = y)) +
  geom_line(col = "white", lwd = 1) +
  geom_vline(aes(xintercept = meanX),
           color = "white", linetype = "dotted",  size = 1) +
  annotate(geom = "segment", x = 2, xend = 2, y = 0,
             yend = dt(2, df = 5),
             color = rgb(0.6769,0.4447,0.7114), lwd = 1.5) +
  labs(y="Density", x = "") +
  coord_cartesian(ylim = c(-0.003, 0.4), clip="off") +
  theme(panel.background = element_rect(fill="grey30", colour="grey30"), 
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(),       
        axis.text.y=element_blank(), axis.ticks.y = element_blank()) +
  scale_x_continuous(breaks=c(RI[2]), limits = c(-3.5,3.5),
                     labels=c("1.96"))
```



```{r, echo = FALSE, cache = TRUE, message = FALSE, warning = FALSE, fig.height=2.5, fig.width=4}
library(ggplot2)
mu = 0
SD = 1
x <- rt(1000000, df = 5)
RI <- c(mu - 2 * SD, mu + 2 * SD)
meanX <- mean(x)

dat <- with(density(x), data.frame(x, y))
ggplot(data = dat, mapping = aes(x = x, y = y)) +
  geom_line(col = "white", lwd = 1) +
  geom_vline(aes(xintercept = meanX),
           color = "white", linetype = "dotted",  size = 1) +
  annotate(geom = "segment", x = -2, xend = -2, y = 0,
             yend = dt(2, df = 5),
             color = rgb(0.6769,0.4447,0.7114), lwd = 1.5) +
  labs(y="Density", x = "") +
  coord_cartesian(ylim = c(-0.003, 0.4), clip="off") +
  theme(panel.background = element_rect(fill="grey30", colour="grey30"), 
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(),         
        axis.text.y=element_blank(), axis.ticks.y = element_blank()) +
  scale_x_continuous(breaks=c(RI[1]), limits = c(-3.5,3.5),
                     labels=c("-1.96"))
```

 \endcol
\endcols



## One sample $z$-test for proportions: Application 

\blue{Critical values}

\begin{tcolorbox}[colback=EMCdark,
                  colframe=EMCdark,
                  width=13cm
                 ]
{\color{white} 

If \textbf{one-tailed}
\newline\newline
critical value$_{\alpha}$:\newline
\texttt{qnorm(p = 0.05, lower.tail = FALSE)}
 \newline\newline
or\newline
 \newline
-critical value$_{\alpha}$: \newline
\texttt{qnorm(p = 0.05, lower.tail = TRUE)}

}

\end{tcolorbox}


## One sample $z$-test for proportions: Application 

\blue{Draw conclusions}

We reject the $H_0$ if: 

- $z$ > critical value$_{\alpha/2}$ or $z$ < - critical value$_{\alpha/2}$

We have -1.43 > -1.96 $\Rightarrow$ we do not reject the $H_0$
\newline

\begincols
  \begincol{0.6\linewidth}

Using ```R``` we obtain the p-value from the $z$-distribution:


```{r, echo = TRUE, cache = TRUE, message = FALSE, warning = FALSE, comment = ""}
2 * pnorm(q = -1.43, lower.tail = TRUE)
```

 \endcol
\begincol{0.4\linewidth}

```{r, echo = FALSE, cache = TRUE, message = FALSE, warning = FALSE, fig.height=2.5, fig.width=4}
library(ggplot2)
mu = 0
SD = 1
x <- rnorm(1000000, mu, SD)
RI <- c(mu - 1.95 * SD, mu + 1.95 * SD)
meanX <- mean(x)

dat <- with(density(x), data.frame(x, y))
ggplot(data = dat, mapping = aes(x = x, y = y)) +
  geom_line(col = "white", lwd = 1) +
  geom_area(mapping = aes(x = ifelse(x >= RI[2] , x, 30)), 
            fill = rgb(0.6769,0.4447,0.7114)) +
  geom_area(mapping = aes(x = ifelse(x <= RI[1] , x, -30)), 
            fill = rgb(0.6769,0.4447,0.7114)) +
  geom_vline(aes(xintercept = meanX),
           color = "white", linetype = "dotted",  size = 1) +
  labs(y="Density", x = "") +
  annotate(geom = "text", x = 0, y = 0.1, label = "85%",
              color = "white", cex = 6) +
  annotate(geom = "text", x = -2.3, y = 0.01, label = "15/2%",
              color = "white", cex = 3) +
  annotate(geom = "text", x = 2.3, y = 0.01, label = "15/2%",
              color = "white", cex = 3) +
  coord_cartesian(ylim = c(-0.003, 0.4), clip="off") +
  theme(panel.background = element_rect(fill="grey30", colour="grey30"), 
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(),         axis.text.y=element_blank(), axis.ticks.y = element_blank()) +
  scale_x_continuous(breaks=c(RI[1], RI[2]), limits = c(-3,3),
                     labels=c("-1.43", "1.43"))
```


 \endcol
\endcols





















 


## Two sample $z$-test for proportions: Theory   

\blue{Scenario} 

Is the probability of being diagnosed with asthma in the Netherlands different than in Belgium?

\blue{Hypothesis}

$H_0: \pi_1 = \pi_2$\newline
$H_1: \pi_1 \neq \pi_2$


## Two sample $z$-test for proportions: Theory  

\blue{Test statistic}

For large sample sizes, the distribution of the test statistic is approximately normal.

\begincols
  \begincol{0.3\textwidth}

Pooled version:

$z = \frac{(p_1-p_2) - 0}{\sqrt{p(1-p)\big(\frac{1}{n_1} + \frac{1}{n_2}\big)}}$

\endcol
\begincol{0.5\linewidth}

Unpooled version:

$z = \frac{(p_1-p_2) - 0}{\sqrt{ \frac{p_1(1-p_1)}{n_1} +  \frac{p_2(1-p_2)}{n_2}  }}$

 \endcol
\endcols

- Sample proportion of group 1: $p_1$
- Sample proportion of group 2: $p_2$
- Number of subjects in group 1: $n_1$
- Number of subjects in group 2: $n_2$
- Total proportion: $p = \frac{n_1p_1 + n_2p_2}{n_1 + n_2}$


## Two sample $z$-test for proportions: Theory  

\blue{Test statistic}

If continuity correction is applied:

\begincols
  \begincol{0.3\textwidth}

Pooled version:

$z = \frac{(p_1-p_2) + \frac{F}{2}\big(\frac{1}{n_1} + \frac{1}{n_2}\big)}{ \sqrt{p(1-p)\big(\frac{1}{n_1} + \frac{1}{n_2}\big)}}$

\endcol
\begincol{0.5\linewidth}

Unpooled version:

$z = \frac{(p_1-p_2) + \frac{F}{2}\big(\frac{1}{n_1} + \frac{1}{n_2}\big)}{\sqrt{ \frac{p_1(1-p_1)}{n_1} +  \frac{p_2(1-p_2)}{n_2}  }}$

 \endcol
\endcols

where\

- $F = -1$ if $p_1 > p_2$
- $F = 1$ if $p_1 < p_2$




## Two sample $z$-test for proportions: Theory  

\blue{Sampling distribution}

- $z$-distribution 
- Critical values and p-value



\blue{Type I error}

- Normally $\alpha = 0.05$

\blue{Draw conclusions}

- Compare test statistic ($z$) with the critical values or the p-value with $\alpha$


## Two sample $z$-test for proportions: Application

\blue{Scenario} 

Is the probability of being diagnosed with asthma in the Netherlands different than in Belgium?

\blue{Hypothesis}

$H_0: \pi_1 = \pi_2$\newline
$H_1: \pi_1 \neq \pi_2$


## Two sample $z$-test for proportions: Application

\begincols
  \begincol{0.4\textwidth}
  
\blue{Collect and visualize data}
\vspace*{-0.5cm}

```{r, echo = FALSE, cache = TRUE, message = FALSE, warning = FALSE}
set.seed(2021)
x1 <- sample(c("No", "Yes"), 100, replace = TRUE)
x2 <- sample(c("No", "Yes"), 100, replace = TRUE)
kable(table(x1), caption = "the Netherlands")
kable(table(x2), caption = "Belgium")
```

\endcol
\begincol{0.65\linewidth}

\blue{Test statistic}

\begin{scriptsize} 
(with no continuity correction and pooled version):
\end{scriptsize}
\newline

$p = \frac{n_1p_1 + n_2p_2}{n_1 + n_2} = \frac{100 \ 0.53 + 100\  0.38}{100 + 100} = 0.46$

$z = \frac{(p_1-p_2) - 0}{\sqrt{p(1-p)\big(\frac{1}{n_1} + \frac{1}{n_2}\big)}} = \frac{0.53-0.38}{\sqrt{0.46(1-0.46) \big( \frac{1}{100} + \frac{1}{100}  \big)}} = 2.13$

\vspace*{1cm}

\blue{Type I error}

$\alpha = 0.05$

 \endcol
\endcols


## Two sample $z$-test for proportions: Application 

\blue{Critical values}

\begincols
  \begincol{0.6\linewidth}

Using ```R``` we get the critical values from the $z$-distribution:\newline
critical value$_{\alpha/2}$ = critical value$_{0.05/2}$

```{r, echo = TRUE, cache = TRUE, message = FALSE, warning = FALSE, comment=""}
qnorm(p = 0.05/2, lower.tail = FALSE)
```

-critical value$_{\alpha/2}$ = -critical value$_{0.05/2}$

```{r, echo = TRUE, cache = TRUE, message = FALSE, warning = FALSE, comment=""}
qnorm(p = 0.05/2, lower.tail = TRUE)
```

 \endcol
\begincol{0.4\linewidth}

\textcolor{white}{texttestestesstdfgzdfgdfgdfgdsdfg}
```{r, echo = FALSE, cache = TRUE, message = FALSE, warning = FALSE, fig.height=2.5, fig.width=4}
library(ggplot2)
mu = 0
SD = 1
x <- rt(1000000, df = 5)
RI <- c(mu - 2 * SD, mu + 2 * SD)
meanX <- mean(x)

dat <- with(density(x), data.frame(x, y))
ggplot(data = dat, mapping = aes(x = x, y = y)) +
  geom_line(col = "white", lwd = 1) +
  geom_vline(aes(xintercept = meanX),
           color = "white", linetype = "dotted",  size = 1) +
  annotate(geom = "segment", x = 2, xend = 2, y = 0,
             yend = dt(2, df = 5),
             color = rgb(0.6769,0.4447,0.7114), lwd = 1.5) +
  labs(y="Density", x = "") +
  coord_cartesian(ylim = c(-0.003, 0.4), clip="off") +
  theme(panel.background = element_rect(fill="grey30", colour="grey30"), 
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(),       
        axis.text.y=element_blank(), axis.ticks.y = element_blank()) +
  scale_x_continuous(breaks=c(RI[2]), limits = c(-3.5,3.5),
                     labels=c("1.96"))
```



```{r, echo = FALSE, cache = TRUE, message = FALSE, warning = FALSE, fig.height=2.5, fig.width=4}
library(ggplot2)
mu = 0
SD = 1
x <- rt(1000000, df = 5)
RI <- c(mu - 2 * SD, mu + 2 * SD)
meanX <- mean(x)

dat <- with(density(x), data.frame(x, y))
ggplot(data = dat, mapping = aes(x = x, y = y)) +
  geom_line(col = "white", lwd = 1) +
  geom_vline(aes(xintercept = meanX),
           color = "white", linetype = "dotted",  size = 1) +
  annotate(geom = "segment", x = -2, xend = -2, y = 0,
             yend = dt(2, df = 5),
             color = rgb(0.6769,0.4447,0.7114), lwd = 1.5) +
  labs(y="Density", x = "") +
  coord_cartesian(ylim = c(-0.003, 0.4), clip="off") +
  theme(panel.background = element_rect(fill="grey30", colour="grey30"), 
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(),         
        axis.text.y=element_blank(), axis.ticks.y = element_blank()) +
  scale_x_continuous(breaks=c(RI[1]), limits = c(-3.5,3.5),
                     labels=c("-1.96"))
```

 \endcol
\endcols


## Two sample $z$-test for proportions: Application 

\blue{Draw conclusions}

We reject the $H_0$ if: 

- $z$ > critical value$_{\alpha/2}$ or $z$ < - critical value$_{\alpha/2}$

We have 2.13 > 1.96 $\Rightarrow$ we reject the $H_0$
\newline

\begincols
  \begincol{0.6\linewidth}
  
Using ```R``` we obtain the p-value from the $z$-distribution:


```{r, echo = TRUE, cache = TRUE, message = FALSE, warning = FALSE, comment = ""}
2 * pnorm(q = 2.13, lower.tail = FALSE)
```

 \endcol
\begincol{0.4\linewidth}

```{r, echo = FALSE, cache = TRUE, message = FALSE, warning = FALSE, fig.height=2.5, fig.width=4}
library(ggplot2)
mu = 0
SD = 1
x <- rnorm(1000000, mu, SD)
RI <- c(mu - 1.95 * SD, mu + 1.95 * SD)
meanX <- mean(x)

dat <- with(density(x), data.frame(x, y))
ggplot(data = dat, mapping = aes(x = x, y = y)) +
  geom_line(col = "white", lwd = 1) +
  geom_area(mapping = aes(x = ifelse(x >= RI[2] , x, 30)), 
            fill = rgb(0.6769,0.4447,0.7114)) +
  geom_area(mapping = aes(x = ifelse(x <= RI[1] , x, -30)), 
            fill = rgb(0.6769,0.4447,0.7114)) +
  geom_vline(aes(xintercept = meanX),
           color = "white", linetype = "dotted",  size = 1) +
  labs(y="Density", x = "") +
  annotate(geom = "text", x = 0, y = 0.1, label = "97%",
              color = "white", cex = 6) +
  annotate(geom = "text", x = -2.3, y = 0.01, label = "3/2%",
              color = "white", cex = 3) +
  annotate(geom = "text", x = 2.3, y = 0.01, label = "3/2%",
              color = "white", cex = 3) +
  coord_cartesian(ylim = c(-0.003, 0.4), clip="off") +
  theme(panel.background = element_rect(fill="grey30", colour="grey30"), 
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(),         axis.text.y=element_blank(), axis.ticks.y = element_blank()) +
  scale_x_continuous(breaks=c(RI[1], RI[2]), limits = c(-3,3),
                     labels=c("-2.13", "2.13"))
```


 \endcol
\endcols

























## Bionomial test: Theory   

\blue{Assumptions}

- Independent observations

\blue{Notes..}

- The binomial test is an exact test
    

## Bionomial test: Theory    

\blue{Scenario} 

Is the probability of being diagnosed with asthma now different than it was 50 years ago?

\blue{Hypothesis}

$H_0: \pi = \pi_0$\newline
$H_1: \pi \neq \pi_0$


## Bionomial test: Theory  

If $n$ is the sample size and $k$ the successes:
$Pr(X = k) = ({n \atop k}) p^k (1-p)^{n-k}$,
where $({n \atop k}) = \frac{n!}{k!(n-k)!}$ and $!$ indicates a factorial


- For any possible outcome of the binomial we obtain the corresponding probability
- We find the p-value by considering the probability of seeing an outcome as, or more, extreme

   - For a one-tailed test, $H_1:\pi<\pi_0$
$p-value = \Pr(X = 0) + \dots + Pr(X = k) = \sum_{i=0}^{k}Pr(X = i) = \sum_{i=0}^{k} ({n \atop i}) p^i (1-p)^{n-i}$ 

   - Calculating a p-value for a two-tailed test is more complicated, since a binomial distribution is not symmetric if $\pi_{0}\neq 0.5$ $\Rightarrow$ we cannot double the p-value from the one-tailed test

\blue{Type I error}

- Normally $\alpha = 0.05$


## Bionomial test: Application

\begin{tcolorbox}[colback=EMC60,
                  colframe=EMC60,
                  width=14cm
                 ]
{\color{white} 

\textbf{Scenario} 

\scriptsize Is the probability of being diagnosed with asthma now lower than it was 50 years ago?

}
\end{tcolorbox}

\begincols
  \begincol{0.4\linewidth}

\vspace*{0.1cm}

\blue{Hypothesis}

$H_0:\pi=\pi_0$\newline
$H_1:\pi < \pi_0$

\vspace*{0.2cm}


\blue{Collect and visualize data}

- $n = 10$
- $k = 3$
- $p = 0.3$
- $\pi_0 = 0.4$ the probability of being diagnosed with asthma 50 years ago


 \endcol
\begincol{0.55\linewidth}

\vspace*{-0.4cm}
\blue{P-value}

$Pr(X <= 3)$
\newline\newline
Using ```R``` we get the p-value:
\vspace*{-0.5cm}
```{r, echo = TRUE, cache = TRUE, message = FALSE, warning = FALSE, comment=""}
pbinom(q = 3, size = 10, prob = 0.4) 
```

\vspace*{0.2cm}
\blue{Draw conclusions}

We do not reject the $H_0$

 \endcol
\endcols


## Bionomial test: Application

\begin{tcolorbox}[colback=EMC60,
                  colframe=EMC60,
                  width=14cm
                 ]
{\color{white} 

\textbf{Scenario} 

\scriptsize Is the probability of being diagnosed with asthma now higher than it was 50 years ago?

}
\end{tcolorbox}

\begincols
  \begincol{0.4\linewidth}

\vspace*{0.1cm}

\blue{Hypothesis}

$H_0:\pi= \pi_0$\newline
$H_1:\pi > \pi_0$

\vspace*{0.2cm}

\blue{Collect and visualize data}

- $n = 10$
- $k = 6$
- $p = 0.6$
- $\pi_0 = 0.4$ the probability of being diagnosed with asthma 50 years ago

 \endcol
\begincol{0.55\linewidth}

\blue{P-value}
$Pr(X >= 6) = 1 - Pr(X < 6) = 1 - Pr(X <= 5)$
\newline\newline
Using ```R``` we get the p-value:
\vspace*{-0.5cm}
```{r, echo = TRUE, cache = TRUE, message = FALSE, warning = FALSE, comment=""}
pbinom(q = 5, size = 10, prob = 0.4, 
       lower.tail = FALSE) 
```

\vspace*{0.2cm}
\blue{Draw conclusions}

We do not reject the $H_0$


 \endcol
\endcols
